PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-05-12 IEEEtran/paper.tex
Merge .tex files, now only use paper.tex to generate both final and annotated version

\DeclareRobustCommand{\add}[1]{#1}
\DeclareRobustCommand{\remove}[1]{}

\DeclareRobustCommand{\addCite}[1]{\textcolor{lightgreen}{\mbox{#1}}}

@@

We present PyTracer, a numerical stability profiling and visualization tool for
Python programs. \pytracer adopts a dynamic approach that evaluates numerical
stability using program execution traces, and is applicable to large programs
without manual intervention. The transparent nature of \pytracer is particularly
valuable in contrast to theoretical approaches based on condition numbers or
backward error analysis~\cite{higham2002accuracy}, which require
detailed modeling of the program and its numerical implementation. Similarly,
static code analysis techniques~\cite{goubault2006static} hardly scale
to large code bases, particularly in dynamically-typed languages such as Python.

\pytracer estimates the variability of floating-point variables by combining
program traces obtained with different numerical perturbations, which requires
(\add{a}\remove{1}) generating numerically-perturbed executions,
(\add{b}\remove{2}) tracing floating-point computations along program
executions, and (\add{c}\remove{3}) visualizing stability evaluations.
\add{PyTracer uses the ``fuzzy" Python environment instrumented with Monte-Carlo
    arithmetic to address the requirement (a). \mbox{\pytracer} addresses the
    remaining requirements with (b) a dynamic instrumentation of Python functions,
    modules, and classes through meta-programming, (c) an interactive Plotly
    dashboard to visualize stability measures throughout program executions.
    \mbox{\pytracer} is available at
    \mbox{\href{https://github.com/big-data-lab-team/pytracer.git}{github.com/big-data-lab-team/pytracer.git}}
    under MIT License.} \remove{\mbox{\pytracer} addresses these requirements with
    (1) a ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic, (2)
    a dynamic instrumentation of Python functions, modules, and classes through
    meta-programming, (3) an interactive Plotly dashboard to visualize stability
    measures throughout program executions. }

@@

% \subsection{Pytracer workflow}
\pytracer adopts the three-step workflow shown in Figure~\ref{fig:workflow}.
First, \pytracer uses \textit{monkey patching} to replace the functions called
by the application with a wrapper that saves their arguments and returned values
in a trace file.
Second, it executes the application multiple times with noise injection, and
aggregates the resulting trace files to compute summary statistics about each
traced variable. Finally, a Plotly dashboard provides interactive visualizations
highlighting unstable code sections.

@@

\pytracer detects numerical instabilities by computing summary statistics across
multiple executions perturbed with numerical noise. While \pytracer's
instrumentation, trace aggregation, and visualization work with various types of
numerical noise, we experimented primarily with Monte-Carlo Arithmetic (MCA) as
it is an accurate model for floating-point errors~\cite{verificarlo}.

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing to simulate
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision. MCA
can be applied in three modes \add{but we only used the} Random Rounding (RR) \add{in our
    experiments} \remove{, Precision Bounding (PB), and full MCA,} which apply
function $inexact$ to the output. \remove{, the inputs, or both, respectively.
    While the RR mode is equivalent to stochastic rounding, the PB mode can also
    identify catastrophic cancellations.} \add{Random Rounding simulates roundoff errors.}

Stochastic arithmetic quantifies numerical error using the number of significant
digits $s$ estimated among the sampled values. A common formula to determine
this number from MCA samples is presented in~\cite{parker1997monte}:
\begin{equation}
    s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a
variable sampled with MCA.
Sohier et al.~\cite{sohier2021confidence} recently provided a generalization of
this formula to include confidence intervals.

We enable MCA in Python programs via Verificarlo~\cite{verificarlo}, a
clang-based compiler~\cite{lattner2008llvm} that replaces floating-point
operations by a generic call to a configurable floating-point model. Several
floating-point models are
available~\cite{chatelain2019automatic,chatelain2019outils}, also called
backends. \pytracer leverages ``fuzzy"~\cite{kiar2020comparing}, a collection of
software tools compiled with Verificarlo. In particular, fuzzy provides
MCA-instrumented versions of the Python interpreter as well as the BLAS, LAPACK,
NumPy, SciPy, and scikit-learn libraries, which enables MCA for a wide range of
existing scientific Python programs. Fuzzy is available in Verificarlo's GitHub
organization at
\href{https://github.com/verificarlo/fuzzy}{\url{github.com/verificarlo/fuzzy}}.

@@

\pytracer stores traces in the pickle format, a binary format to serialize
Python objects. The main advantage of the pickle format compared to other
serialization approaches such as marshal or JSON is the ability to serialize
most Python objects with a native compression. In addition, pickle writes
Python objects sequentially, which preserves the temporal ordering required by
subsequent trace analyses.

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information are loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. Therefore, it assumes that the application is
single-threaded and that the control flow does not depend on a random state.
\add{Therefore, \mbox{\pytracer} cannot merge traces resulting from a different
    control flow path and stops the parsing when encountering one.}

@@

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA in the Python interpreter \add{as well as} BLAS, LAPACK, NumPy,
SciPy, and scikit-learn. \add{Only instrumented the Python interpreter would
    lead to only perturbe pure Python floating-point operations and thus to miss
    operations in compiled codes (i.e. Cython and shared libraries).} We used the default virtual
precision in Verificarlo, namely 24 bits for single-precision floating-point
values and 53 bits for double-precision values, which simulates machine error.
We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.

\add{Monte Carlo Arithmetic samples the floating-point variability of a given
    program, and hence the more MCA repetitions we have, the more accurate will be
    the sampling distribution. In theory, an infinite number of MCA repetitions
    should give the exact distribution. While in practice low sample size (3$\simeq$30)
    can give a good estimate, the issue is to capture "rare events" that could make the
    computation strongly diverge from the mean. For now, determining the number of
    MCA samples to use is still an open research question. 
    Recent theoretical works}~\cite{sohier2021confidence}
\add{on the significant digits computation allow computing an associated
    confidence interval depending on the sample size and the underlying distribution of the outputs.}

@@

\subsection{\pytracer allows visualizing numerical stability at a fine grained level}

\add{We illustrate on two examples how PyTracer is effective for analyzing
    the numerical stability at a fine grained level: Face recognition and Bayesian Ridge Regression.}
The face recognition example uses eigenfaces (eigenvectors) and Support Vector
Machines (SVMs) to classify faces from the Labeled Faces in the Wild dataset.
This example uses a Principal Component Analysis (PCA) to extract the eigenfaces
and then trains an SVM classifier to predict the name associated with a face.
\add{When executing within fuzzy environment, this example crashed with the following
    Python error raised: \texttt{"SVD did not converge"}.}
This example is the only one to raise an error due to the
non-convergence of the Singular Value Decomposition (SVD) performed in the PCA.
The PCA uses the Randomized SVD~\cite{halko2011finding} method
that first computes a low-rank approximation of the input matrix using a
stochastic algorithm and then uses this approximation as input for the SVD. 
\add{When walking through the code with PyTracer, we see that}
the \texttt{randomized\_svd} scikit-learn function uses the Divide \& Conquer SVD
method \texttt{gesdd}. The LAPACK library has two main methods for computing the SVD:
\texttt{gesdd} and \texttt{gesvd}. The former uses a Divide \& Conquer approach,
while the latter uses a QR decomposition. While both methods are expected to
have the same accuracy~\cite{nakatsukasa2013stable}, the Divide \& Conquer
makes the SVD to not converge with MCA.

Replacing \texttt{gesdd} for \texttt{gesvd} solves the SVD convergence issue.
\pytracer allows to quickly visualize the new precision of the result.
Figure~\ref{fig:randomized_svd_S} shows the precision of the eigenvalues
computed displayed in decreasing order with \texttt{gesvd} \add{in single precision}, 
which varies between 15 and 20
significant bits out of the 24 available in single precision. The classifier
metrics (precision, recall, F1-score, and support) are identical to the IEEE
execution for all MCA samples. \add{We can note that even with using the \texttt{gesvd},
    the precision of the singular values (vectors) is not uniform. 
    Indeed, we can see that the precision of the lowest singular values (vectors) decreases 
    until reaching 0 significant bits.}

The second example, compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
non-significant. \pytracer traces reveal that the SVD solver used is the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer reveals an issue in the singular values computed.}
\add{We can see on Figure}~\ref{fig:BRR_svd} \add{that the last singular values is
    \texttt{NaN} due to numerical error during the SVD although the remaining
    singular values exhibit high numerical precision (between 44 and 50 bits).
    Surprinsingly, this \texttt{NaN} value does not raise issue and is silently
    propagated. Indeed, the stopping criterion compares that the sum of the absolute
    differences between coefficients of two successive iterations but coefficients
    contain \texttt{NaN} since they are computed from singular values. Hence, the
    following right condition is never met \texttt{if iter\_!=0 and np.sum(np.abs(coef\_old\_-coef\_))<self.tol:} since the comparison with a
    \texttt{NaN} is always false.}

\add{The OLS uses the Least Square to fit the model. PyTracer shows that the
    coefficients fitted are \texttt{NaN} values. By looking at the callgraph, we can
    see that the scipy calls the \texttt{lstsq} method from the
    \texttt{scipy.linalg} package which uses the \texttt{gelsd} LAPACK function.
    Figure}~\ref{fig:OLS_svd} \add{shows the singular values returned by the
    \texttt{lstsq} method shows that one singular value has a \texttt{NaN} value
    that corrupted the calculation of the coefficients. By using the \texttt{gelss},
    PyTracer shows that the \texttt{NaN} value disapeared from the singular values
    (fig.}~\ref{fig:OLS_svd_fixed}\add{) and thus does not contaminated the computation of
    the coefficients that exhibit a precision of 42 bits in average. We can see that
    the full access to all values computed by the application with PyTracer allows
    the user to quickly identify the issue.}

In conclusion, \pytracer and the MCA model were helpful to detect the
instability of the \texttt{gesdd} method mentioned in several GitHub issues,
including the LAPACK repository. Furthermore, it is worth mentioning that a Pull
Request has recently been submitted to let the user select the SVD method and
avoid this kind of issue. Thus, \pytracer can detect instabilities that affect
fundamental scientific computing components such as the LAPACK library. \add{The
    unify view of all traces allows the user to quickly navigate through the
    different numerical precision of variables which saves time for the developer
    compare to the print statement debugging.}
% The Pytracer findings can therefore be extended to many application domains. %
% \tristan{conclusion needs to be discussed. Note to self: generaliser un peu la
% conclusion: pytracer permet de se rendre compte de problemes de stabilite dont
% plusieurs librairies sont dependantes.}

@@

\begin{figure}
    \centering
    \begin{subfigure}{0.49\linewidth}
        \includegraphics[width=\linewidth]{figure/BRR_singular_values.png}
        \caption{BRR}
        \label{fig:BRR_svd}
    \end{subfigure}
    \begin{subfigure}{0.49\linewidth}
        \includegraphics[width=\linewidth]{figure/OLS_singular_values.png}
        \caption{OLS}
        \label{fig:OLS_svd}
    \end{subfigure}
    \begin{subfigure}{0.49\linewidth}
        \includegraphics[width=\linewidth]{figure/BRR_singular_values_fixed.png}
        \caption{BRR (fixed)}
        \label{fig:BRR_svd_fixed}
    \end{subfigure}
    \begin{subfigure}{0.49\linewidth}
        \includegraphics[width=\linewidth]{figure/OLS_singular_values_fixed.png}
        \caption{OLS (fixed)}
        \label{fig:OLS_svd_fixed}
    \end{subfigure}
    \caption{\add{Numerical precision of the singular values of the SVD from the
            Bayesian Ridge model fitting (left) and from the Ordinary Least Square model
            fitting (right) in the Bayesian Ridge Regression example. Red star
            represents \texttt{NaN} value. Both SVD give \texttt{NaN} coefficients.
            Switching from Divide \& Conquer to QR factorization improve numerical precision.
            PyTracer helps to quickly visualize improvements.}}
\end{figure}

@@

\pytracer's tracing overhead consists of (1) module instrumentation time
(initialization overhead) and (2) data write time during the execution (runtime
overhead). The initialization overhead is proportional to the number of
functions and classes traced. For example, we measured an average initialization
time of 8 seconds during our experiments with  NumPy, SciPy, and Scikit-learn.
This overhead can be reduced by excluding intensively used functions such as
\texttt{numpy.array} from the instrumentation.
Figure~\ref{fig:performance_tracing} summarizes \pytracer's tracing overhead for
SciPy and scikit-learn examples. \textit{Slowdown} represents the slowdown of
\pytracer without fuzzy measured as the ratio between the actual execution time
(without \pytracer) and the execution time using \pytracer. The \textit{Slowdown
    RR} \remove{and \textit{Slowdown MCA}} shows the slowdown within the fuzzy environment.
Without fuzzy, \pytracer introduces an average slowdown of $17\times$ high for
the examples running less than 8 seconds because \pytracer spends most of its
time in the initialization step. Since this step is a constant factor of the
number of functions instrumented, we also measured the instrumentation cost
itself (\textit{Slowdown (Instrumentation)}) calculated by subtracting the
initialization time to \textit{Slowdown}. We then obtain an average slowdown of
$3.5\times$. Finally, the \remove{fuzzy instrumentation} \add{using MCA
    operations (throuhg fuzzy) instead of the native ones} increases the average
slowdown to $37\times$ for RR\remove{and $43\times$ for Full MCA}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/performance.pdf}
    \caption{
        \pytracer tracing overheads. \textit{Slowdown} represents the slowdown
        of \pytracer without fuzzy measured as the ratio between the actual
        execution time (without \pytracer) and the execution time using
        \pytracer. \textit{Slowdown (Instrumentation)} show the actual
        instrumentation cost computed by subtracting the initialization step.
        The \textit{Slowdown (RR)} \remove{and \textit{Slowdown (MCA)} are} \add{is} the slowdown
        of \pytracer using the fuzzy environment with \add{the} RR \remove{and Full MCA} mode\remove{s}. The
        first \remove{first} \add{last} four examples measure the initialization
        steps depending on the modules instrumented. 
        % \tristan{there's too much text in this caption, move some of it to the text, the caption should just explain the difference between the 4 slowdowns. From the text I'm not sure what is the difference between slowdown and slowdown(RR)/slowdown(MCA).}
    }
    \label{fig:performance_tracing}
\end{figure}

@@

\clearpage
\appendix

\include{annex}

>>>>

\usepackage{ifthen}
\newif\ifANNOTATED
\ifthenelse{\equal{\detokenize{annotated}}{\jobname}}{
\ANNOTATEDtrue
}{
\ANNOTATEDfalse
}

\ifANNOTATED
\DeclareRobustCommand{\add}[1]{\sethlcolor{lightgreen}\hl{#1}}
\DeclareRobustCommand{\remove}[1]{\textcolor{lightred}{\sout{#1}}}
\DeclareRobustCommand{\addCite}[1]{\textcolor{lightgreen}{\mbox{#1}}}
\newcommand{\TG}[1]{\color{red}\textsc{From Tristan}: #1\xspace\color{black}}
\else
\DeclareRobustCommand{\add}[1]{#1}
\DeclareRobustCommand{\remove}[1]{}
\DeclareRobustCommand{\addCite}[1]{#1}
\newcommand{\TG}[1]{}
\fi

@@

We present PyTracer, a numerical stability profiling and visualization tool for
Python programs. \pytracer adopts a dynamic approach that evaluates numerical
stability using program execution traces, and is applicable to large programs
without manual intervention. The transparent nature of \pytracer is particularly
valuable in contrast to theoretical approaches based on condition numbers or
backward error analysis~\addCite{\cite{higham2002accuracy}}, which require
detailed modeling of the program and its numerical implementation. Similarly,
static code analysis techniques~\addCite{\cite{goubault2006static}} hardly scale
to large code bases, particularly in dynamically-typed languages such as Python.

\pytracer estimates the variability of floating-point variables by combining
program traces obtained with different numerical perturbations, which requires
(\add{a}\remove{1}) generating numerically-perturbed executions,
(\add{b}\remove{2}) tracing floating-point computations along program
executions, and (\add{c}\remove{3}) visualizing stability evaluations.
\add{PyTracer addresses (a) by relying on the ``fuzzy" Python environment instrumented with Monte-Carlo
    arithmetic, (b) by dynamically instrumenting Python functions,
    modules, and classes through meta-programming, and (c) by visualizing stability measures throughout program executions
    with an interactive Plotly
    dashboard.
    \mbox{\pytracer} is available at
    \mbox{\href{https://github.com/big-data-lab-team/pytracer.git}{github.com/big-data-lab-team/pytracer.git}}
    under MIT License.} \remove{\mbox{\pytracer} addresses these requirements with
    (1) a ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic, (2)
    a dynamic instrumentation of Python functions, modules, and classes through
    meta-programming, (3) an interactive Plotly dashboard to visualize stability
    measures throughout program executions. }

@@

\TG{I'm still unsure what to do with this part, it looks out of place.}
% \add{PyTracer follows a three steps approach:}
% \begin{enumerate}
%     \item \add{Automatic instrumentation of Python programs to save traces containing input and output values of all functions.}
%     \item \add{Traces aggregation to produce a statistics summary.}
%     \item \add{Interactive visualization to navigate through code execution.}
% \end{enumerate}

\pytracer adopts the three-step workflow shown in Figure~\ref{fig:workflow}.
First, \pytracer uses \textit{monkey patching} to replace the functions called
by the application with a wrapper that saves their arguments and returned values
in a trace file.
Second, it executes the application multiple times with noise injection, and
aggregates the resulting trace files to compute summary statistics about each
traced variable. Finally, a Plotly dashboard provides interactive visualizations
highlighting unstable code sections.

\add{
    The main goal of PyTracer is to help developpers visualize the numerical
    stability of Python applications. PyTracer saves time for the user by
    instrumenting and aggregating values. In the same way that GDB makes the user
    save time when debugging an application by providing all information in one
    layout, PyTracer allows the user to visualize the overall numerical quality to
    quickly detect instabilities. The PyTracer visualization layout improves result
    analysis in any case since 1) the user does not have to manually instrument its
    code to dump raw data and creates specific post-processing scripts to visualize
    it 2) PyTracer allows the user to export the raw data traced to extend the
    analysis. 
}

\add{
    The main challenge addressed by PyTracer is the transparent instrumentation of Python 
    applications. Indeed, since Python is an interpreted language, one has no complete 
    view of the function that will be executed by a given program. 
    Moreover, for Python applications used in HPC environments, one has
    most of the time no access to source codes or with limited write access.
    Therefore, PyTracer tackles this issue by instrumenting the function dynamically 
    thanks to the Python introspection.
}

\add{
    The trace-based approach used by PyTracer permits to limit the 
    overhead at runtime since the statistics computation is reported to the postprocess analysis.
    However, it limits the aggregation to deterministic executions,
    although control-flow analysis can help to merge divergent execution paths.
    Hence, PyTracer is restricted to single-threaded applications even if its usage
    is limited in Python due to the Global Interperter Lock (GIL) mechanism
    that allows only one thread to hold the control of the Python interpreter.
}

% \subsection{Pytracer workflow}

@@

\pytracer detects numerical instabilities by computing summary statistics across
multiple executions perturbed with numerical noise. While \pytracer's
instrumentation, trace aggregation, and visualization work with various types of
numerical noise, we experimented primarily with Monte-Carlo Arithmetic (MCA) as
it is an accurate model for floating-point errors~\addCite{\cite{verificarlo}}.

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing to simulate
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
\add{In our experiments, we used MCA in Random Rounding (RR) mode, which
    simulates roundoff errors by applying function $inexact$ only to the
    output returned by a function. The Precision Bounding (PB) mode can
    identify other types of instabilities, in particular catastrophic
    cancellations, however, it is more difficult to apply to large code bases
    as it often leads to execution errors that require human intervention.}

Stochastic arithmetic quantifies numerical error using the number of significant
digits $s$ estimated among the sampled values. A common formula to determine
this number from MCA samples is presented in~\cite{parker1997monte}:
\begin{equation}
    s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a
variable sampled with MCA. \add{This can be seen as the ratio signal over noise.}
Sohier et al.~\cite{sohier2021confidence} recently provided a generalization of
this formula to include confidence intervals.

We enabled MCA in Python programs via Verificarlo~\cite{verificarlo}, a
clang-based compiler~\cite{lattner2008llvm} that replaces floating-point
operations by a generic call to a configurable floating-point model. Several
floating-point models are
available~\cite{chatelain2019automatic,chatelain2019outils}, also called
backends. \pytracer leverages ``fuzzy"~\cite{kiar2020comparing}, a collection of
software tools compiled with Verificarlo. In particular, fuzzy provides
MCA-instrumented versions of the Python interpreter as well as the BLAS, LAPACK,
NumPy, SciPy, and scikit-learn libraries, which enables MCA for a wide range of
existing scientific Python programs. Fuzzy is available in Verificarlo's GitHub
organization at
\href{https://github.com/verificarlo/fuzzy}{\url{github.com/verificarlo/fuzzy}}.

\add{Monte Carlo Arithmetic samples the floating-point variability of a given
    program, and hence the more MCA repetitions we have, the more accurate the
    distribution will be. In theory, an infinite number of MCA repetitions
    should give the exact distribution. While in practice low sample sizes (3 to 30)
    can give good estimates, the issue is to capture ``rare events" that could make the
    computation strongly diverge from the mean. Determining the number of
    MCA samples to use for a given instrumented code is still an open research question. 
    Recent theoretical works}~\cite{sohier2021confidence}
\add{on the estimation of significant digits allow computing an associated
    confidence interval depending on the sample size and the underlying distribution.}

@@

\pytracer stores traces in the pickle format, a binary format to serialize
Python objects. The main advantages of the pickle format compared to other
serialization approaches such as marshal or JSON are its ability to serialize
most Python objects and its native compression.  In addition, pickle writes
Python objects sequentially, which preserves the temporal ordering required by
subsequent trace analyses.

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information are loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. Therefore, it assumes that the application is
single-threaded and that the control flow does not depend on a random state.
\add{Therefore, \mbox{\pytracer} cannot merge traces resulting from different
    control flow paths and will stop the parsing when encountering such a situation.}

@@

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA in the Python interpreter \add{as well as} BLAS, LAPACK, NumPy,
SciPy, and scikit-learn. \add{Instrumenting only the Python interpreter would
    lead to only perturb pure Python floating-point operations and thus miss
    operations in compiled codes (i.e., Cython and shared libraries).} \add{We
    used MCA in Random Rounding (RR) mode, with 5 repetitions}. We used the default virtual
precision in Verificarlo, namely 24 bits for single-precision floating-point
values and 53 bits for double-precision values, which simulates machine error.
We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.

@@

\subsection{\pytracer allows visualizing numerical stability at a fine-grained level}

\add{We illustrate on two examples how PyTracer is effective for analyzing
    numerical stability in complex codes at a fine-grained level: Face recognition and Bayesian Ridge Regression.}
The face recognition example uses eigenfaces (eigenvectors) and Support Vector
Machines (SVMs) to classify faces from the Labeled Faces in the Wild dataset.
This example uses a Principal Component Analysis (PCA) to extract the eigenfaces
and then trains an SVM classifier to predict the label associated with a face.
\add{When executing in PyTracer, this example crashed because the Singular Value Decomposition (SVD) did not converge.}
This example is the only one to raise an error due to the
non-convergence of the SVD performed in the PCA.
The PCA uses the Randomized SVD~\cite{halko2011finding} method
that first computes a low-rank approximation of the input matrix using a
stochastic algorithm and then uses this approximation as input for the SVD. 
\add{When walking through the code with PyTracer, we see that}
the \texttt{randomized\_svd} scikit-learn function uses the Divide \& Conquer SVD
method \texttt{gesdd}. The LAPACK library has two main methods for computing the SVD:
\texttt{gesdd} and \texttt{gesvd}. The former uses a Divide \& Conquer approach,
while the latter uses a QR decomposition. While both methods are expected to
have the same accuracy~\cite{nakatsukasa2013stable}, the Divide \& Conquer
does not converge with MCA.

Replacing \texttt{gesdd} for \texttt{gesvd} solved the SVD convergence issue.
\pytracer allows to quickly visualize the new precision of the result.
Figure~\ref{fig:face_recognition_svd} shows the precision of the eigenvalues
computed displayed in decreasing order with \texttt{gesvd} \add{in single precision}, 
which varies between 15 and 20
significant bits out of the 24 available in single precision. The classifier
metrics (precision, recall, F1-score, and support) are identical to the IEEE
execution for all MCA samples. \add{We can note that even with the \texttt{gesvd} method,
    the precision of the singular values and eigenvectors is not uniform. 
    Indeed, we can see that the precision of the lowest singular values and eigenvectors decreases 
    until reaching 0 significant bits.}

The second example compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
non-significant. \pytracer traces reveal that the SVD solver used is once again the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer shows an issue in the singular values computed.}
\add{We can see in Figure}~\ref{fig:BRR_svd} \add{that the last singular values is
    \texttt{NaN} due to numerical errors in the SVD, although the remaining
    singular values exhibit high numerical precision (between 44 and 50 bits).
    Surprinsingly, this \texttt{NaN} value does not raise an exception and is silently
    propagated. Indeed, the stopping criterion is never verified and the 
    optimization stops after reaching the maximal number of iterations.
    Figure}~\ref{fig:BRR_svd_fixed} \add{shows that uses
    the \texttt{gesvd} instead of the \texttt{gesdd} method removes the \texttt{NaN} value
    although the last singular values stills have a low precision (below 0).
    Following Equation}~\ref{eq:sig-digits}, \add{a number of significant digits below 0 means that the
    variability ($\sigma$) is higher than the average value ($\mu$)}.

\add{Similarly, for OLS, PyTracer tracers show that the
    regression coefficients fitted are \texttt{NaN} values. By looking at the callgraph, we can
    see that the scipy calls the \texttt{lstsq} method from the
    \texttt{scipy.linalg} package which uses the \texttt{gesdd} LAPACK function.
    Figure}~\ref{fig:OLS_svd} \add{shows that one singular value is a \texttt{NaN} value,
    which corrupted the calculation of the coefficients. When using the \texttt{gesvd} method, 
    the \texttt{NaN} value disapears 
    (Fig.}~\ref{fig:OLS_svd_fixed}\add{), resulting in a precision of
    42 bits in average in the coefficients. We can see that having
    access to all the values computed by the application with PyTracer allows
    the user to efficiently identify numerical issues.}

In conclusion, \pytracer and the MCA model were helpful to detect the
instability of the \texttt{gesdd} method mentioned in several GitHub issues,
including the LAPACK repository. Furthermore, it is worth mentioning that a Pull
Request has recently been submitted to let the user select the SVD method and
avoid this kind of issue. Thus, \pytracer can detect instabilities that affect
fundamental scientific computing components such as the LAPACK library. \add{The
    unified view of all traces allows the user to quickly navigate through the
    numerical precision of different variables.}
% The Pytracer findings can therefore be extended to many application domains. %
% \tristan{conclusion needs to be discussed. Note to self: generaliser un peu la
% conclusion: pytracer permet de se rendre compte de problemes de stabilite dont
% plusieurs librairies sont dependantes.}

@@

\begin{figure}
    \centering
    \begin{subfigure}{0.49\linewidth}
        \includegraphics[width=\linewidth]{figure/BRR_singular_values.png}
        \caption{BRR (\texttt{gesdd})}
        \label{fig:BRR_svd}
    \end{subfigure}
    \begin{subfigure}{0.49\linewidth}
        \includegraphics[width=\linewidth]{figure/OLS_singular_values.png}
        \caption{OLS (\texttt{gesdd})}
        \label{fig:OLS_svd}
    \end{subfigure}
    \begin{subfigure}{0.49\linewidth}
        \includegraphics[width=\linewidth]{figure/BRR_singular_values_fixed.png}
        \caption{BRR (\texttt{gesvd})}
        \label{fig:BRR_svd_fixed}
    \end{subfigure}
    \begin{subfigure}{0.49\linewidth}
        \includegraphics[width=\linewidth]{figure/OLS_singular_values_fixed.png}
        \caption{OLS (\texttt{gesvd})}
        \label{fig:OLS_svd_fixed}
    \end{subfigure}
    \caption{\add{Numerical precision of the singular values of the SVD from the
            Bayesian Ridge model fitting (left) and from the Ordinary Least Square model
            fitting (right) in the Bayesian Ridge Regression example. Red stars
            represent \texttt{NaN} values. 
            Switching from Divide \& Conquer (\texttt{gesdd}) to QR (\texttt{gesvd}) factorization improves numerical precision.
            PyTracer helps visualizing improvements.}}
\end{figure}

@@

\pytracer's tracing overhead consists of (1) module instrumentation time
(initialization overhead) and (2) data write time during the execution (runtime
overhead). The initialization overhead is proportional to the number of
functions and classes traced. For example, we measured an average initialization
time of 8 seconds during our experiments with  NumPy, SciPy, and Scikit-learn.
This overhead can be reduced by excluding intensively used functions such as
\texttt{numpy.array} from the instrumentation.
Figure~\ref{fig:performance_tracing} summarizes \pytracer's tracing overhead for
SciPy and scikit-learn examples. \textit{Slowdown} represents the slowdown of
\pytracer without fuzzy measured as the ratio between the actual execution time
(without \pytracer) and the execution time using \pytracer. The \textit{Slowdown
    RR} \remove{and \textit{Slowdown MCA}} shows the slowdown within the fuzzy environment.
Without fuzzy, \pytracer introduces an average slowdown of $17\times$ for
the examples running less than 8 seconds because \pytracer spends most of its
time in the initialization step. Since this step is a constant factor of the
number of functions instrumented, we also measured the instrumentation cost
itself (\textit{Slowdown (Instrumentation)}) calculated by subtracting the
initialization time to \textit{Slowdown}. We then obtain an average slowdown of
$3.5\times$. Finally, the \remove{fuzzy instrumentation} \add{using MCA
    operations (through fuzzy) instead of the native ones} increases the average
slowdown to $37\times$ for RR\remove{and $43\times$ for Full MCA}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/performance.pdf}
    \caption{
        \pytracer tracing overheads. \textit{Slowdown} represents the slowdown
        of \pytracer without fuzzy measured as the ratio between the actual
        execution time (without \pytracer) and the execution time using
        \pytracer. \textit{Slowdown (Instrumentation)} show the actual
        instrumentation cost computed by subtracting the initialization step.
        The \textit{Slowdown (RR)} \remove{and \textit{Slowdown (MCA)} are} \add{is} the slowdown
        of \pytracer using the fuzzy environment with \add{the} RR \remove{and Full MCA} mode\remove{s}. The
        \remove{first} \add{last} four examples measure the initialization
        steps depending on the modules instrumented. 
        % \tristan{there's too much text in this caption, move some of it to the text, the caption should just explain the difference between the 4 slowdowns. From the text I'm not sure what is the difference between slowdown and slowdown(RR)/slowdown(MCA).}
    }
    \label{fig:performance_tracing}
\end{figure}

@@

% \appendix
% \include{annex}
